Copyright (c) 2022, Emily Ricotta
All rights reserved.

This source code is licensed under the BSD-style license found in the LICENSE file in the root directory of this source tree. 

---
title: "PERSIST study - Creation and application of 4 paramater logistic model"
author: "Emily Ricotta"
date: "2022-12-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(magrittr)
require(drc)
require(dr4pl)

set.seed(1234)
```

```{r load data, include=FALSE}
#Load provided demo data
plate1<-load("plate1_demodata.RData")
```

Some of the following code is from https://katatrepsis.com/2018/09/24/elisa-analysis-in-r/

The functions conc2od and od2conc were created using the equations in https://static-content.springer.com/esm/art%3A10.1038%2Fs41467-020-20383-x/MediaObjects/41467_2020_20383_MOESM1_ESM.pdf

Concentration calculations solved using dr4pl package

```{r concentration/optical density equation functions}
#Equation A in paper's supplementary figure 10
conc2od = function(conc, slope, lower, upper, ic50) {
  setNames(lower + (conc ^ slope * (upper - lower)) / (conc^slope + ic50 ^ slope), "OD")
}

#Equation B in paper's supplementary figure 10
od2conc = function(od, slope, lower, upper, ic50) {
  if(od <= lower) {
    0
  } else if(od >= upper) {
    1/400
  } else {
    setNames(ic50 * ((lower - upper) / (od - upper) - 1)^(1/upper), "conc")
  }
}
```

```{r sIgG concentraton function}
sIgG_model<-function(x){
  #Select analyte for plot data
  fps<-forplot %>%
  filter(analyte=="Spike IgG" & !is.na(concentration)) %>%
  distinct()
  #Construct 4 parameter logistic regression model 
  model.sIgG <- dr4pl(dose = fps$concentration,
                      response = fps$OD,
                      method.init = "logistic")
  #Name the coefficients from the model something intuitive
  coefs <- setNames(
    model.sIgG$parameters,
    c("Upper", "EC50", "Slope", "Lower"))
  #Calculate the IC50
  IC50<-IC(model.sIgG, 50)    
  OD_at_IC50<-conc2od(IC50, coefs["Slope"], coefs["Lower"], coefs["Upper"], coefs["EC50"])
  #Plot the model
  plot(model.sIgG)+
    geom_hline(yintercept = OD_at_IC50) +
    geom_vline(xintercept=IC50)
  #Choose the optical density closest to the IC50 calculated from the standard curve
  sIgG_conc <- x %>%
    filter(analyte=="Spike IgG") %>%
    group_by(sampid) %>%
    mutate(ODave_IC50 = nth(ODave, which.min(abs(ODave-OD_at_IC50))))
  #Keep only the row that is closest to the IC50. If there is a tie, pick the highest dilution
  sIgG_conc %<>%
    group_by(sampid) %>%
    mutate(ODneat=case_when(OD_dilution=="Neat"~ODave,
                            TRUE~NA_real_),
           ODneat=zoo::na.locf(ODneat)) %>%
    filter(ODave==ODave_IC50) %>%
    filter(row_number()==n()) %>%
    mutate(multiplier=case_when(OD_dilution=="Neat"~10,
                                OD_dilution=="1:10 Dilution"~100,
                                OD_dilution=="1:100 Dilution"~1000,
                                OD_dilution=="1:1000 Dilution"~10000,
                                TRUE~NA_real_))
  #Calculate the concentration from the standard curve
  sIgG_conc %<>%
    filter(analyte=="Spike IgG") %>%
    mutate(OD_flag=case_when(ODave <= coefs["Lower"] ~ -1,
                             ODave >= coefs["Upper"] ~ 1,
                             TRUE ~ 0),
           conc=od2conc(ODave, coefs["Slope"], coefs["Lower"], coefs["Upper"], coefs["EC50"]),
           conc.ug_ml=(conc*multiplier)*1000)
  sIgG_conc %<>%
    group_by(sampid) %>%
    mutate(cutoff=case_when(ODneat>0.674720844 &
                                 conc.ug_ml>0 ~ 1,
                               TRUE~0),
           cutoff=max(cutoff))
    sIgG_conc %<>%
    dplyr::select(c(sampid, analyte, ODave, OD_flag, OD_dilution, multiplier, conc, conc.ug_ml, cutoff))
}
```

```{r Plate 1 concentration calculations}
#Create the dataset
forplot1<-plate1 %>%
  dplyr::select(c("analyte", "OD"=Blank1_ODave, "dilution"=Curve1_dilution))
forplot2<-plate1 %>%
  dplyr::select(c("analyte", "OD"=Blank2_ODave, "dilution"=Curve2_dilution))
forplot<-bind_rows(forplot1, forplot2)

#Calculate the concentration based on the dilution factor of the well
forplot %<>%
  mutate(concentration=case_when(
    dilution=="1:100"~(1/100)*0.0025,
    dilution=="1:300"~(1/300)*0.0025,
    dilution=="1:500"~(1/500)*0.0025,
    dilution=="1:600"~(1/600)*0.0025,
    dilution=="1:1000"~(1/1000)*0.0025,
    dilution=="1:1200"~(1/1200)*0.0025,
    dilution=="1:2500"~(1/2500)*0.0025,
    dilution=="1:2400"~(1/2400)*0.0025,
    dilution=="1:5000"~(1/5000)*0.0025,
    dilution=="1:4800"~(1/4800)*0.0025,
    dilution=="1:10,000"~(1/10000)*0.0025,
    dilution=="1:9600"~(1/9600)*0.0025,
    TRUE~NA_real_))

sIgG_conc<-sIgG_model(plate1)

```
